Embedding Intuitionistic into Classical Logic

### 2022-10-12 FoSSaCS/fossacs-EIICL.tex
fossacs first final draft

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic which allows us to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We then transfer this procedure to first-order logic which allows us to give a embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

We shall now give an overview of our main results and highlight the key arguments. Recall that our goal is to give an effective translation procedure that, for a given formula $\varphi$, yields a formula $\varphi^\#$, such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning:
It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$.
Still we propose a similar transformation.
A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity.

\begin{theorem}\label{thm:Tseytin}
For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
$\hspace{1cm} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
\end{theorem}

The main transformation then proceeds in three steps:
%
1) We encode as a first-order sentence that the considered formula holds for every Kripke frame.
Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory this step is rather straightforward.
%
2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
%
3) We can then argue that if there is a counter-model for the resulting formula then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke frame.

Summarizing, we obtain the following result:

\begin{theorem}
\label{thm:reduction-propositional}
	Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
	\begin{itemize}
		\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
		\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occur in $k$.
	\end{itemize}
Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Futhermore, there is an effective procedure for translating counter-models between the sentences.
\end{theorem}

Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
We can show that this QBF is satisfiable if and only if there exists a counter-model for our original formula:

\begin{theorem}
	Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$. There is an effective procedure that produces a QBF $\varphi^Q$ of size $\mathcal O(|\mathcal S|\cdot|\mathcal F_\to| + |\mathcal F_\to|^3)$ with $2\cdot |\mathcal F_\to|-1$ quantifier alternations such that $\varphi$ is intuitionistically valid if and only if $\varphi^Q$ is not satisfiable. For fixed $N\in\mathbb N$, deciding intuitionistic validity for formulas $\bigwedge \mathcal S\to P$ where $\mathcal S$ is as above and $|\mathcal F_\to| = N$ is therefore in $\Sigma_{2N-1}$.
\end{theorem}

%\begin{theorem}
%	For every predicate formula $\varphi$ there exists an effective procedure that yields an atom $P$ and a set $\mathcal S$ of formulas of the forms
%	$$\forall \vec z(A(\vec a)\to (B(\vec b)\wedge C(\vec c))), \forall \vec z((A(\vec a)\wedge B(\vec b))\to C(\vec c)),$$$$\forall \vec z(A(\vec a)\to (B(\vec b)\vee C(\vec c))),
%	\forall \vec z((A(\vec a)\vee B(\vec b))\to C(\vec c)),$$$$ \forall \vec z((A(\vec a)\to B(\vec b))\to C(\vec c)),\forall \vec z(\forall xA(\vec a)\to B(\vec b)),$$$$ \forall \vec z(A(\vec a)\to\forall xB(b)), \forall \vec z(\exists xA(\vec a)\to B(\vec b)), \forall \vec z(A(\vec a)\to\exists xB(b))$$for atomic $A, B, C$ and vectors of variables $\vec z, \vec a, \vec b, \vec c$ such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The size of $\mathcal S$ and time complexity of the transformation are linear in the input size.
%\end{theorem}

We then move to the first-order case. Here the construction is a bit more involved but the underlying approach is the same.
Again we first perform a Tseytin-like transformation.
Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
The details of this translation are given in Section~\ref{section:tseytin}.
We then proceed similarly to the propositional case.
Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame but on the other hand also proper domain elements.
We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world.
The main transformation then again proceeds in three steps:
Step 1) and 2) are analogous to the propositional case.
The real difference arises in step 3):
While we can also establish the existence of a canonic counter-model whose frame only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
This is not surprising since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models.
Therefore we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics.
However since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable.
%
Summarizing, we obtain the following result:

\begin{theorem}
\label{thm:reduction-first-order-short}
There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $\varphi$, however, for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$, and $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore, there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
\end{theorem}
A more detailed account of Theorem~\ref{thm:reduction-first-order-short} is given in Theorem~\ref{thm:fo-translation}.

These results provide an effective framework for using classical provers to check intuitionistic validity. First experiments with the Vampire theorem prover~\cite{kovacs2013first} show promise.

@@

\begin{corollary}
	If a formula in normal form as in Lemma~\ref{lemma:propositional-normal-form} in intuitionistically invalid then there exists a Kripke counter-model which is a rooted tree with height and degree bounded by $|\mathcal X|$.
\end{corollary}

@@

\begin{theorem}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implication $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to(C^\lambda)\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $P^\#$ as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$ encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

\section{IPC to QBF}

@@

A simple counting argument shows that for each $n < |\mathcal X|$ we have $|\text{Valid}(n)|\in \mathcal O(|\mathcal X|^2)$, $|\text{Persistent}(n)|, |\text{Sat}_{\mathcal R}(n)|\in \mathcal O(|\varphi|)$ and $|\text{Sat}_{\mathcal X}(n)|\in \mathcal O(|\mathcal X|)$ so overall we get:
\begin{lemma}
	The size of $\varphi^Q$ is in $\mathcal O(|\varphi|\cdot|\mathcal X| + |\mathcal X|^3)$.
\end{lemma}

@@

We now give an analogous transformation for first order logic. Our trick of lifting the sentence to first-order logic to encode the Kripke semantics no longer works in such a straightforward manner, since we already are in first-order logic. However we do it nonetheless! The domain of our classical model will feature on the one hand elements representing worlds in a Kripke frame and on the other hand the domain-elements from the Kripke model. We reconcile these notions by introducing a special binary predicate $E$, first considered in~\cite{baaz2006skolemization} and expanded onto in~\cite{iemhoff2010eskolemization}, where $E(x, u)$ encodes that $x$ is an element of the domain of the world $u$. To encode that some $n$-ary relation $A$ holds at a world $u$ we replace each $n$-ary relation symbol $A$ with a $n+1$-ary relation symbol $A$, interpreting $A(\vec x, u)$ as "$A(\vec x)$ holds at u". Furthermore we write $\vec E(\vec x, u)$ for $\bigwedge\{E(x_i, u)\:|\:x_i\in \vec x\}$. With this we obtain the following encoding of Kripke Semantics:

\begin{definition}
	Let $\varphi$ be a propositional formula. Define $\varphi^{u}$ inductively:
	\begin{itemize}
		\item $\bot^u = \bot$.
		\item $A(\vec x)^{u} = A(\vec x, u)$ for every relation symbol $A$.
		\item $(\varphi\circ\psi)^u = \varphi^u\circ\psi^u$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^u = \forall w(u\preceq w\to\varphi^{w}\to\psi^{w})$
		\item $(\exists x\varphi)^u = \exists x(E(x,u)\wedge\varphi^u)$
		\item $(\forall x\varphi)^u = \forall w(u\preceq w\to E(x, w)\to \varphi^w)$
	\end{itemize}
	where $w$ is some new variable. Let $K(\varphi)$ encode the theory of Kripke structures as well as the predicate $E$, i.e.
	\begin{align*}
		K(\varphi) = \:& \text{PartialOrder}(\preceq) \wedge \forall u \forall w (u\preceq w\to \text{DomainSubset}(u, w)) \wedge\\
		& \forall u(\text{World(u)}\to \text{DomainClosed}(u))\wedge \forall u\forall w (u\preceq w\to \text{Persistent}(u, w)).
	\end{align*}
	with e.g.:
	\begin{align*}
		\text{PartialOrder}(\preceq) = &\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{DomainSubset}(u, w) = &\:\forall x(E(x, u)\to E(x, w))\\
		\text{World}(u) = &\:\exists xE(x, u)\\
		\text{DomainClosed}(u) = &\:\bigwedge\left\{\forall\vec x(\vec E(\vec x, u)\to E(f(\vec x), u))\:|\:\text{$f$ a function in $\varphi$}\right\}\\
		\text{Persistent}(u, w) = &\:\bigwedge\left\{\forall\vec x(\vec A(\vec x, u)\to A(\vec x, w))\:|\:\text{$A$ a relation in $\varphi$}\right\}
	\end{align*}
	Then define
	$$\varphi^C = K(\varphi)\to\forall u(\text{World}(u)\to \varphi^u).$$
\end{definition}

@@

Since we are mixing worlds and proper intuitionistic domain elements in a single classical domain there are some details to be taken care of. A proof is given in the appendix as Lemma~\ref{proof:fo-simplification}.

@@

\begin{wrapfigure}[6]{r}{0.4\textwidth}
		\vspace*{-1.7cm}
		\begin{tabular}{l|c|c}
			Embedding&&\\
			 Embedding (idem.)&557&20.9\%\\
			ileanCoP 1.2&&\\
			nanoCoP-i 2.0&858&32.1\%\\\hline
			Total&2669&100\%
		\end{tabular}
		\caption{Solved problems.}
		\label{fig:a}
\end{wrapfigure}

We have benchmarked our first-order translation on version 1.1.2 of the ILTP problem set~\cite{iltp} by performing our translation and then examining the generated classical problem using Vampire~\cite{kovacs2013first}. For comparison the state-of-the-art intuitionistic provers ileanCoP 1.2~\cite{otten2008leancop} and nanoCoP-i 2.0~\cite{otten2021nanocop} are used. All benchmarks were conducted on an Intel Core i5-8400 2.80GHz machine running Linux with a time limit of 10s per problem. Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results can be seen in Figure \ref{fig:a}. In general our approach falls short of the state of the art. However when binning for $|\mathcal Q|$ and $|\mathcal X|$ as seen in Figure~\ref{fig:b} we observe something interesting.
\begin{figure}
	\centering
	\begin{tabular}{c|c|c|c|c|c|c}
		$|\mathcal Q|$&$|\mathcal X|$&Embedding&Embedding (idem.)&ileanCoP&nanoCoP-i 2.0&Total\\\hline
		0&0&&157&&61&178\\
		0&1-2&&130&&96&400\\
		0&3-5&&17&&40&105\\
		0&6-10&&4&&7&19\\
		0&11-20&&0&&3&34\\
		0&21+&&7&&1&17\\
		1-2&0&&14&&15&17\\
		1-2&1-2&&67&&206&270\\
		1-2&3-5&&77&&122&312\\
		1-2&6-10&&15&&28&64\\
		1-2&11+&&4&&6&12\\
		3-5&0&&14&&16&16\\
		3-5&1-2&&11&&36&49\\
		3-5&3-5&&15&&58&159\\
		3-5&6-10&&16&&83&199\\
		3-5&11+&&2&&3&38\\
		6+&0-2&&1&&50&50\\
		6+&3+&&3&&24&605
	\end{tabular}
	\caption{Solved problems binned by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}
Leveraging the power of Vampire, for low $|\mathcal X|$ and $|\mathcal Q|$ the embedding performs on par or even outperforms the dedicated intuitionistic provers. For larger values the state of the art provers have an advantage, in particular when both parameters are combined. Therefore our tool is already useful if it is known that $|\mathcal X|$ and $|\mathcal Q|$ are small. Another hopeful sign is that the inclusion of idempotence increases performance in particular for larger $|\mathcal X|$ and $|\mathcal Q|$. Further improvements could be achieved by introducing more axioms based on structural properties of the formula.
Another interesting observation is that $|\mathcal X|$ and $|\mathcal Q|$ also seem to correlate with success rate for the dedicated intuitionistic provers.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic which allows us to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We then transfer this procedure to first-order logic which allows us to give a embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

We shall now give an overview of our main results and highlight the key arguments. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

The general idea is to use the Kripke semantics for intuitionistic logic, which are expressed in classical logic, to perform an embedding of formulas. That is, for each formula $\varphi$ we give a formula $\varphi^\#$ that is classically valid if and only if $\varphi$ is intuitionistically valid, establishing a dual to the double negation translation.

The key property that allows us to reduce our encoding, which a priori targets first order logic, to propositional logic, is a reduction of counter-models, which is a proper dual to validity for intuitionistic logic. In particular, given $$\varphi = \left(\bigwedge\mathcal R\wedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form} we establish:

\begin{lemma}
	If $\varphi$ is intuitionistically invalid then there exists a Kripke counter-model which is a rooted tree with height and degree bounded by $|\mathcal X|$.
\end{lemma}

This allows us to embed intuitionistic propositional logic into classical propositional logic.

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a formula $\varphi^\#$ with $\mathcal |\varphi^\#|\in\mathcal O(|\mathcal R|\cdot\exp(|\mathcal X|\log(|\mathcal X|)))$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
\end{lemma}

The exact details can be found in Theorem~\ref{thm:prop-translation}. The essential mechanisms is the elimination $\forall$-quantifiers by enumeration of worlds in a sufficiently general Kripke Frame. Instead of expressing correctness at each world of the model we can express that the worlds along some path in the tree are correct. The branching of paths can be expressed through quantification in a quantified boolean formula (QBF) which gives us the following:

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a QBF $\varphi^Q$ with $\mathcal |\varphi^Q|\in\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$ such that $\varphi$ is intuitionistically invalid if and only if $\varphi^Q$ is a satisfiable QBF.
\end{lemma}

Again, the details can be found in Section~\ref{sec:qbf}. Next we attempt to transport these results to the first order case. There are formulas for which only counter-examples with an infinite Kripke Frame exist. Therefore, there is no hope of completely eliminating terms representing worlds from the final embedding. This also means that our language necessarily becomes more complex. However a number of reductions is still possible.

\begin{theorem}
	\label{thm:reduction-first-order-short}
	There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $\varphi$, however, for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$, and $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore, there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
\end{theorem}

A detailed account of the translation can be found in definition~\ref{def:fo-translation}. Finally, we benchmark our translation by translating all problems in the ILTP problem set~\cite{iltp} and applying the Vampire theorem prover~\cite{kovacs2013first} to the translated instances. While this approach falls short of the absolute state of the art in general, under the right circumstances it is a reasonable approach. Since there is still a lot of room for improvement this is a hopeful first sign.

@@

\begin{corollary}
	If a formula in normal form as in Lemma~\ref{lemma:propositional-normal-form} is intuitionistically invalid then there exists a Kripke counter-model which is a rooted tree with height and degree bounded by $|\mathcal X|$.
\end{corollary}

@@

\begin{theorem}\label{thm:prop-translation}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implication $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to(C^\lambda)\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $P^\#$ as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$ encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

\section{IPC to QBF}\label{sec:qbf}

@@

A simple counting argument shows that for each $n < |\mathcal X|$ we have $|\text{Valid}(n)|\in \mathcal O(|\mathcal X|^2)$, $|\text{Persistent}(n)|, |\text{Sat}_{\mathcal R}(n)|\in \mathcal O(|\mathcal R|)$ and $|\text{Sat}_{\mathcal X}(n)|\in \mathcal O(|\mathcal X|)$ so overall we get:
\begin{lemma}
	The size of $\varphi^Q$ is in $\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$.
\end{lemma}

@@

We now give an analogous transformation for first order logic. Our trick of lifting the sentence to first-order logic to encode the Kripke semantics no longer works in such a straightforward manner, since we already are in first-order logic. However we do it nonetheless! The domain of our classical model will feature on the one hand elements representing worlds in a Kripke frame and on the other hand the domain-elements from the Kripke model. We reconcile these notions by introducing a special binary predicate $E$, first considered in~\cite{baaz2006skolemization} and expanded onto in~\cite{iemhoff2010eskolemization}, where $E(x, u)$ encodes that $x$ is an element of the domain of the world $u$. To encode that some $n$-ary relation $A$ holds at a world $u$ we extend each $n$-ary relation symbol to a $n+1$-ary relation symbol of the same name, interpreting $A(\vec x, u)$ as "$A(\vec x)$ holds at u". Furthermore we write $\vec E(\vec x, u)$ for $\bigwedge\{E(x_i, u)\:|\:x_i\in \vec x\}$. With this we obtain the following encoding of Kripke Semantics:

\begin{definition}\label{def:fo-translation}
	Let $\varphi$ be a propositional formula. Define $\varphi^{u}$ inductively:
	\begin{itemize}
		\item $\bot^u = \bot$.
		\item $A(\vec x)^{u} = A(\vec x, u)$ for every relation symbol $A$.
		\item $(\varphi\circ\psi)^u = \varphi^u\circ\psi^u$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^u = \forall w(u\preceq w\to\varphi^{w}\to\psi^{w})$
		\item $(\exists x\varphi)^u = \exists x(E(x,u)\wedge\varphi^u)$
		\item $(\forall x\varphi)^u = \forall x\forall w(u\preceq w\to E(x, w)\to \varphi^w)$
	\end{itemize}
	where $w$ is some new variable. Let $K(\varphi)$ encode the theory of Kripke structures as well as the predicate $E$, i.e.
	\begin{align*}
		K(\varphi) = \:& \text{PartialOrder}(\preceq) \wedge \forall u \forall w (u\preceq w\to \text{DomainSubset}(u, w)) \wedge\\
		& \forall u(\text{World(u)}\to \text{DomainClosed}(u))\wedge \forall u\forall w (u\preceq w\to \text{Persistent}(u, w)).
	\end{align*}
	with e.g.:
	\begin{align*}
		\text{PartialOrder}(\preceq) = &\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{DomainSubset}(u, w) = &\:\forall x(E(x, u)\to E(x, w))\\
		\text{World}(u) = &\:\exists xE(x, u)\\
		\text{DomainClosed}(u) = &\:\bigwedge\left\{\forall\vec x(\vec E(\vec x, u)\to E(f(\vec x), u))\:|\:\text{$f$ a function in $\varphi$}\right\}\\
		\text{Persistent}(u, w) = &\:\bigwedge\left\{\forall\vec x(\vec A(\vec x, u)\to A(\vec x, w))\:|\:\text{$A$ a relation in $\varphi$}\right\}
	\end{align*}
	Then define
	$$\varphi^C = K(\varphi)\to\forall u(\text{World}(u)\to \varphi^u).$$
\end{definition}

@@

\begin{proof}
	Since we are mixing worlds and proper intuitionistic domain elements in a single classical domain there are some details to be taken care of. We proceed by translation of counter-models.

	Suppose first we have a counter-model $\mathcal M = (M, I)$ to $\varphi^C$. As a Kripke frame $(W, \leq)$ take $ W = \{m\in M\:|\:\text{ World}^I(m)\}$ and let $\leq$ be $\preceq^I$ restricted to $W$. Then let $M_u = \{m\in M\:|\: E(m, u)\}$ and let $f^{I_u}$ be $f^I$ restricted to $M_u$ and $A^{I_u}(\vec x): \Leftrightarrow A(\vec x, u)$. It is then a straightforward check of definitions that this defines a Kripke counter-model to $\varphi$.

	The other direction is a bit more involved. Suppose we have a Kripke counter-model to $\varphi$ with frame $(W, \preceq)$ and family of $\Sigma$-structures $(M_w, I_w)_{w\in W}$. In particular, since it is a counter-model there exists $w_0\in W$ and a variable assignment $v$ with $w_0, v\not\models\varphi$. Let $W_0 = \{w\in W\:|\: w_0\preceq w\}$ and define an equivalence relation $\sim$ on $\{(x, u)\:|\:u\in W_0, x\in M_u\}$ via $(x, u)\sim (y, w)$ iff $x = y$ and there exists $v\in W_0$ comparable with both $u, w$ such that $x\in v$ and denote the equivalence class of $(x, u)$ with $[x, u]$. Let $M = W_0\cup \{[x, u]\:|\:u\in W_0, x\in M_u\}$.
	Now have
	\begin{itemize}
		\item $\preceq^I = \leq$ when restricted to $W_0$ and $\prec$ can be interpreted as an arbitrary PO otherwise.
		\item $E^I(m, w)$ iff $w\in W_0$ and $m = [x, w]$ for some $x\in M_w$.
		\item $f^I(m_1\dots m_n) =\begin{cases}
			f^{I_u}(x_1\dots x_n), &\substack{\text{if there are $ u\in W_0, x_i\in M_u$}\\ \text{with $m_i= [x_i, u]$ for all $i$,}}\\
			w_0, & \text{otherwise.}
		\end{cases}$
		\item ${A}^I(m_1\dots m_n, u) \Leftrightarrow\begin{cases}
			A^{I_u}(x_1\dots x_n), &\substack{\text{if $u\in W_0$ and $\exists x_i\in M_u$}\\\text{with $m_i= [x_i, u]$ for all $i$,}}\\
			\top, & \text{otherwise.}
		\end{cases}$
	\end{itemize}
	One easily verifies that these are well-defined. Let us now briefly check that this indeed defines a counter-model. Clearly World$(w_0)$ holds and $K(\varphi)$ also holds. All that remains to show is that $M, I, v[w_0/u]\not\models\varphi^u$. We proceed by induction on the height of $\varphi$. Consider e.g. the case where $\varphi$ is of the form $\forall x\psi$.
	We have to show that
	$$M, I, v[w_0/u]\models \forall x\forall w(u\preceq w\to E(x, w)\to \psi^w)$$
	does not hold for some $v$. Now since $\varphi$ is intuitionistically invalid there exists $w_1\in W_0$ and $m\in M_{w_1}$ such that $w_1, v[m/x]\not\models\psi$. By induction hypothesis $M, I, v[w_1/w]\not\models\psi^w$ for some $v$. By definition $M, I, v[w_1/w, [m, w_1]/x]\models E(x, w)$ and $w_0\preceq^I w_1$ holds. Therefore indeed $$M, I, v[w_0/u, w_1/w, [m, w_1]/x]\not\models \forall w(u\preceq w\to E(x, w)\to \psi^w)$$ holds and we are done. The other cases are analogous.\hfill$\square$
\end{proof}

@@

\begin{wrapfigure}[6]{r}{0.4\textwidth}
		\vspace*{-1.7cm}
		\begin{tabular}{l|c|c}
			Embedding&571&21.4\%\\
			 Embedding (idem.)&557&20.9\%\\
			ileanCoP 1.2&875&32.8\%\\
			nanoCoP-i 2.0&858&32.1\%\\\hline
			Total&2669&100\%
		\end{tabular}
		\caption{Solved problems.}
		\label{fig:a}
\end{wrapfigure}
